---
title: "MSBA Captstone -Modeling"
author: "Maddie Lee"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
execute:
  warning: false
  message: false
---
```{r include=FALSE, echo = FALSE, warnings = FALSE, message = FALSE}
library(MASS)
library(tidyverse)
library(knitr)
library(readxl)
library(Metrics)
library(car)
library(caret)
library(glmnet)
library(forecast)
library(prophet)
library(randomForest)
library(ggplot2)
library(caret)
library(tidymodels)
library(xgboost)
library(timetk)
```

# Introduction

## Business Problem Statement

Swire Coca-Cola (SCCU) aims to optimize delivery logistics by shifting low-volume customers to cost-efficient Alternate Routes to Market (ARTM/ white
truck delivery) using third-party services. However, this may inadvertently move high-growth potential customers to less personalized service, risking
revenue loss and weaker relationships. The purpose of this project is to establish a reliable, data-informed systematic method to identify and predict
high-potential customers, balancing efficiency with sustainable growth.

## Discussion of Modeling

A supervised machine learning model will be used to predict total annual delivery cost per customer. By leveraging historical sales and delivery data, it will help identify ARTM customers with potential for volume growth, ensuring cost-efficient truck assignments and forecasting high-cost customers before they exceed thresholds. The model is trained using one year of data as the training set and the next year as the test set, with key features such as order volume, delivery success rate, and trade channel. Performance will be evaluated based on key metrics such as RMSE, R², and MAPE to evaluate the model's accuracy. 

Once the calculations and predictions are complete, unsupervised learning (clustering) can be used to segment deliveries into White Truck vs. Red Truck categories based on key logistics variables. 

A dashboard will then be constructed to illustrate results. The goal of this dashboard is to help logistics teams and operations managers classify deliveries into White Truck vs. Red Truck based on key order attributes. By integrating clustering techniques, this dashboard will automate truck assignments, reduce delivery costs, and improve operational efficiency. Ultimately, the desire is to determine how to integrate the insights into a routine strategy to support long-term growth while maintaining logistical efficiency.  

# Modeling

## Preprocessing

### Loading & Joining Data

Loading in data sets:

```{r}
# Import data sets
setwd("/Users/madelinelee/Downloads/")
transactional_data <- read.csv(file = "transactional_data.csv", stringsAsFactors = TRUE)
customer_profile <- read.csv(file = "customer_profile.csv", stringsAsFactors = TRUE)
customer_address_and_zip_mapping <- read.csv(file = "customer_address_and_zip_mapping.csv", stringsAsFactors = TRUE)
delivery_cost_data <- read_excel("delivery_cost_data.xlsx")
```

Combining transaction and customer data into one data set:

```{r}
# Join transactional and customer profile data
transaction_customer_data <- transactional_data %>%
                             left_join(customer_profile, by = c("CUSTOMER_NUMBER" = "CUSTOMER_NUMBER"))

# Join customer address data
transaction_customer_data <- transaction_customer_data %>%
                             left_join(customer_address_and_zip_mapping, by = c("ZIP_CODE" = "zip"))
```

### Feature Engineering from the EDA

Creating new variables from Exploratory Data Analysis (EDA):

**Total Units Ordered**

 - Description: used to quantify total units ordered, which are then used to determine when a threshold is reached
 - Calculation: sum of two existing variables

```{r}
# Total Units Ordered
transaction_customer_data$TOTAL_UNITS_ORDERED <- transaction_customer_data$ORDERED_GALLONS + transaction_customer_data$ORDERED_CASES
```

**Total Units Loaded**

 - Description: used to quantify total units loaded
 - Calculation: sum of two existing variables

```{r}
# Total Units Loaded
transaction_customer_data$TOTAL_UNITS_LOADED <- transaction_customer_data$LOADED_GALLONS + transaction_customer_data$LOADED_CASES
```

**Total Units Delivered**

 - Description: used to quantify total units delivered
 - Calculation: sum of two existing variables

```{r}
# Total Units Loaded
transaction_customer_data$TOTAL_UNITS_DELIVERED <- transaction_customer_data$DELIVERED_GALLONS + transaction_customer_data$DELIVERED_CASES
```

**Order Grouping**

 - Description: used to identify customers who are ordering less or more than 400 units per year
 - Calculation: logical statement utilizing level of detail (LOD) calculation functionality

```{r}
transaction_customer_data$ORDER_GROUPING <- (transaction_customer_data %>%
                                            group_by(YEAR, CUSTOMER_NUMBER) %>%
                                            mutate(TOTAL_UNITS_ORDERED = sum(TOTAL_UNITS_ORDERED)) %>%
                                            ungroup() %>%
                                            mutate(ORDER_GROUPING = as.factor(ifelse(TOTAL_UNITS_ORDERED < 400, 
                                                                                                'Customers Ordering Less than 400', 
                                                                                                'Customers Ordering More than 400'))))$ORDER_GROUPING
```

**Customer Group Number (Rollup)**

- Description: used to aggregate transactions up to the customer group number, if desired
- Calculation: logical statement using two existing variables

```{r}
transaction_customer_data$CUSTOMER_GROUP_NUMBER_ROLLUP <- ifelse(!is.na(transaction_customer_data$PRIMARY_GROUP_NUMBER), 
                                                                 transaction_customer_data$PRIMARY_GROUP_NUMBER, 
                                                                 transaction_customer_data$CUSTOMER_NUMBER)
```

### New Feature Engineering: Calculating Delivery Cost per Customer

Steps provided by SWIRE:

 1. Determine the volume range for each category (Fountain and Bottles/ Cans)
 2. Retrieve the median delivery cost per unit from the data set.
 3. Multiply the cost per unit by the quantity purchased for each category.
 4. Sum the delivery costs to get the total.
 
We'll start by calculating the total volume purchased by each customer in each category per year. 

```{r}
# Summarize the total volume purchased for each category
annual_volume <- transaction_customer_data %>%
  group_by(YEAR, CUSTOMER_NUMBER, COLD_DRINK_CHANNEL) %>%
  summarize(
    Total_Fountain = sum(ORDERED_GALLONS),
    Total_Bottles_Cans = sum(ORDERED_CASES)
  ) %>%
  ungroup()

# View the result
head(annual_volume, n = 5)
tail(annual_volume, n = 5)
```

Next, we need to determine the range each fountain and bottle/ can value falls into, writing a function to do so.

```{r}
# Ensure Vol Range is a character
delivery_costs <- delivery_cost_data %>%
  mutate(`Vol Range` = as.character(`Vol Range`))

# Function to get the delivery cost based on volume range
get_delivery_cost <- function(volume, category, channel) {
  if (volume >= 0 & volume <= 149) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "0 - 149") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 150 & volume <= 299) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "150 - 299") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 300 & volume <= 449) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "300 - 449") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 450 & volume <= 599) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "450 - 599") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 600 & volume <= 749) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "600 - 749") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 750 & volume <= 899) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "750 - 899") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 900 & volume <= 1049) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "900 - 1049") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 1050 & volume <= 1199) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "1050 - 1199") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 1200 & volume <= 1349) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "1200 - 1349") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else if (volume >= 1350) {
    cost <- delivery_costs %>%
      filter(`Cold Drink Channel` == channel, `Applicable To` == category, `Vol Range` == "1350+") %>%
      select(`Median Delivery Cost`) %>%
      pull()
  } else {
    cost <- 0
  }
  return(ifelse(length(cost) > 0, cost, 0))
}
```

Next, we'll calculate the total delivery cost for each customer.

```{r}
# Calculate the total delivery cost for each customer
annual_volume <- annual_volume %>%
  rowwise() %>%
  mutate(
    Fountain_Cost = get_delivery_cost(Total_Fountain, "Fountain", COLD_DRINK_CHANNEL) * Total_Fountain,
    Bottles_Cans_Cost = get_delivery_cost(Total_Bottles_Cans, "Bottles and Cans", COLD_DRINK_CHANNEL) * Total_Bottles_Cans,
    Total_Delivery_Cost = Fountain_Cost + Bottles_Cans_Cost
  ) %>%
  ungroup()


```

Last, we'll add the total delivery cost column to our main data set.

```{r}
# Merge the total delivery cost back into the transaction_customer_data table
transaction_customer_data <- transaction_customer_data %>%
  left_join(annual_volume %>% select(CUSTOMER_NUMBER, YEAR, Total_Delivery_Cost), 
            by = c("CUSTOMER_NUMBER" = "CUSTOMER_NUMBER", "YEAR" = "YEAR"))

# View the result
head(transaction_customer_data, n = 5)
tail(transaction_customer_data, n = 5)
```

## Defining Train & Test Sets

Because some customers have data in just 2023 or 2024, we want to make sure we split those customers into train and test sets proportionately.

The plan is to:

 - Identify customers with data in 2023, 2024, both years, and only one of the years.
 - Split the customers into train and test sets based on the specified proportions.
 - Create the train and test sets using the transaction_customer_data table.
 
```{r}
# Identify customers with data in 2023 and 2024
customers_2023 <- transaction_customer_data %>%
  filter(YEAR == 2023) %>%
  select(CUSTOMER_NUMBER) %>%
  distinct()

customers_2024 <- transaction_customer_data %>%
  filter(YEAR == 2024) %>%
  select(CUSTOMER_NUMBER) %>%
  distinct()

# Identify customers with data in both years, only 2023, and only 2024
customers_both_years <- intersect(customers_2023$CUSTOMER_NUMBER, customers_2024$CUSTOMER_NUMBER)
customers_only_2023 <- setdiff(customers_2023$CUSTOMER_NUMBER, customers_2024$CUSTOMER_NUMBER)
customers_only_2024 <- setdiff(customers_2024$CUSTOMER_NUMBER, customers_2023$CUSTOMER_NUMBER)

# Function to split customers into train and test sets
split_customers <- function(customers, train_ratio = 0.7) {
  set.seed(123) # For reproducibility
  train_indices <- sample(seq_len(length(customers)), size = floor(train_ratio * length(customers)))
  train_customers <- customers[train_indices]
  test_customers <- customers[-train_indices]
  return(list(train = train_customers, test = test_customers))
}

# Split customers into train and test sets
split_both_years <- split_customers(customers_both_years)
split_only_2023 <- split_customers(customers_only_2023)
split_only_2024 <- split_customers(customers_only_2024)

# Combine train and test customers
train_customers <- c(split_both_years$train, split_only_2023$train, split_only_2024$train)
test_customers <- c(split_both_years$test, split_only_2023$test, split_only_2024$test)

# Create train and test sets
train_set <- transaction_customer_data %>%
  filter(CUSTOMER_NUMBER %in% train_customers)

test_set <- transaction_customer_data %>%
  filter(CUSTOMER_NUMBER %in% test_customers)

# View the results
print("Train Set:")
head(train_set, n = 5)
tail(train_set, n = 5)
print("Test Set:")
head(test_set, n = 5)
tail(test_set, n = 5)
```

Let's test to make sure each customer (and all of their transactions) end up in either the train set or the test set, and aren't split among the two:

```{r}
# Check for overlap
overlap_customers <- intersect(train_customers, test_customers)
if (length(overlap_customers) == 0) {
  print("No overlap between train and test customers.")
} else {
  print("Overlap found between train and test customers.")
}
```

These sets are still quite large let's aggregate each row up to the customer and year level so there is one row per customer per year. This will decrease the time required to create and test the models.

```{r}
# Assigning columns a group
dimensions <- c("ORDER_TYPE", 
                "PRIMARY_GROUP_NUMBER",
                "FREQUENT_ORDER_TYPE",
                "FIRST_DELIVERY_DATE",
                "ON_BOARDING_DATE",
                "COLD_DRINK_CHANNEL",
                "TRADE_CHANNEL",
                "SUB_TRADE_CHANNEL",
                "LOCAL_MARKET_PARTNER",
                "CO2_CUSTOMER",
                "ZIP_CODE",
                "full.address",
                "ORDER_GROUPING",
                "CUSTOMER_GROUP_NUMBER_ROLLUP")

measures <- c("ORDERED_CASES", 
              "LOADED_CASES", 
              "DELIVERED_CASES",
              "ORDERED_GALLONS",
              "LOADED_GALLONS",
              "DELIVERED_GALLONS",
              "TOTAL_UNITS_ORDERED",
              "TOTAL_UNITS_LOADED",
              "TOTAL_UNITS_DELIVERED") 

drop_columns <- c("WEEK", 
                  "TRANSACTION_DATE") 

already_aggregated <- "Total_Delivery_Cost" 

# Aggregate the data
aggregated_data <- transaction_customer_data %>%
  select(-one_of(drop_columns)) %>% # Drop unnecessary columns
  group_by(CUSTOMER_NUMBER, YEAR) %>%
  summarize(across(all_of(dimensions), first), # Keep dimensions as-is
            across(all_of(measures), sum, .names = "sum_{col}"), # Sum measures
            across(all_of(already_aggregated), first)) # Keep already aggregated column as-is

# View the results
print("Aggregated Data:")
head(aggregated_data, n = 5)
tail(aggregated_data, n = 5)

# Aggregate the train set
aggregated_train_set <- train_set %>%
  select(-one_of(drop_columns)) %>% # Drop unnecessary columns
  group_by(CUSTOMER_NUMBER, YEAR) %>%
  summarize(across(all_of(dimensions), first), # Keep dimensions as-is
            across(all_of(measures), sum, .names = "sum_{col}"), # Sum measures
            across(all_of(already_aggregated), first)) # Keep already aggregated column as-is

# Aggregate the test set
aggregated_test_set <- test_set %>%
  select(-one_of(drop_columns)) %>% # Drop unnecessary columns
  group_by(CUSTOMER_NUMBER, YEAR) %>%
  summarize(across(all_of(dimensions), first), # Keep dimensions as-is
            across(all_of(measures), sum, .names = "sum_{col}"), # Sum measures
            across(all_of(already_aggregated), first)) # Keep already aggregated column as-is

# View the results
print("Aggregated Train Set:")
head(aggregated_train_set, n = 5)
tail(aggregated_train_set, n = 5)

print("Aggregated Test Set:")
head(aggregated_test_set, n = 5)
tail(aggregated_test_set, n = 5)
```

Lastly, I want to make sure that I am using 2023 data to predict Total Units ordered in 2024.

```{r}
# Separate the data into 2023 and 2024 subsets
train_data_2023 <- aggregated_train_set %>% filter(YEAR == 2023)
train_data_2024 <- aggregated_train_set %>% filter(YEAR == 2024)
test_data_2023 <- aggregated_test_set %>% filter(YEAR == 2023)
test_data_2024 <- aggregated_test_set %>% filter(YEAR == 2024)

# Merge the 2023 and 2024 data based on customer number
train_merged_data <- merge(train_data_2023, train_data_2024, by = "CUSTOMER_NUMBER", suffixes = c("_2023", "_2024"))
test_merged_data <- merge(test_data_2023, test_data_2024, by = "CUSTOMER_NUMBER", suffixes = c("_2023", "_2024"))

# Select relevant
train_set <- train_merged_data %>% select(CUSTOMER_NUMBER, ends_with("_2023"), sum_TOTAL_UNITS_ORDERED_2024)
test_set <- test_merged_data %>% select(CUSTOMER_NUMBER, ends_with("_2023"), sum_TOTAL_UNITS_ORDERED_2024)
```

# Modeling

```{r modeling}
options(warn=-1) 
# Define predictors and target variable
predictor_vars <- grep("_2023$", colnames(train_set), value = TRUE)
target_var <- "sum_TOTAL_UNITS_ORDERED_2024"

# Remove rows with missing values
train_set <- train_set %>% drop_na(all_of(c(predictor_vars, target_var)))
test_set <- test_set %>% drop_na(all_of(predictor_vars))

# Impute missing values (if needed)
train_set[predictor_vars] <- train_set[predictor_vars] %>% mutate_all(~ ifelse(is.na(.), median(., na.rm = TRUE), .))
test_set[predictor_vars] <- test_set[predictor_vars] %>% mutate_all(~ ifelse(is.na(.), median(., na.rm = TRUE), .))

# Define cross-validation setup
train_control <- trainControl(method = "cv", number = 5)

# Train models (Fix: Keep y in train_set)
models <- list()

models$lm <- train(
  as.formula(paste(target_var, "~ .")),
  data = train_set,
  method = "lm",
  trControl = train_control
)

models$ridge <- train(
  as.formula(paste(target_var, "~ .")),
  data = train_set,
  method = "glmnet",
  trControl = train_control,
  tuneGrid = expand.grid(alpha = 0, lambda = seq(0.001, 1, length = 10))
)

models$lasso <- train(
  as.formula(paste(target_var, "~ .")),
  data = train_set,
  method = "glmnet",
  trControl = train_control,
  tuneGrid = expand.grid(alpha = 1, lambda = seq(0.001, 1, length = 10))
)

models$rf <- train(
  as.formula(paste(target_var, "~ .")),
  data = train_set,
  method = "rf",
  trControl = train_control,
  tuneGrid = expand.grid(mtry = c(2, 5, 10))
)

xgb_grid <- expand.grid(
  nrounds = c(50, 100, 200), eta = c(0.01, 0.1, 0.3),
  max_depth = c(3, 6, 9), gamma = 0, colsample_bytree = 0.8,
  min_child_weight = 1, subsample = 0.8
)

models$xgb <- train(
  as.formula(paste(target_var, "~ .")),
  data = train_set,
  method = "xgbTree",
  trControl = train_control,
  tuneGrid = xgb_grid
)

# Elastic Net Regression (combines Ridge & Lasso)
enet_grid <- expand.grid(alpha = seq(0, 1, 0.1), lambda = seq(0.001, 1, length = 10))
models$enet <- train(as.formula(paste(target_var, "~ .")), data = train_set, method = "glmnet", 
                     trControl = train_control, tuneGrid = enet_grid)

# Decision Tree Model
models$dt <- train(
  as.formula(paste(target_var, "~ .")),
  data = train_set,
  method = "rpart",
  trControl = train_control,
  tuneGrid = expand.grid(cp = seq(0.001, 0.1, by = 0.01))
)

# Gradient Boosting Machine (GBM)
models$gbm <- train(
  as.formula(paste(target_var, "~ .")),
  data = train_set,
  method = "gbm",
  trControl = train_control,
  verbose = FALSE
)


# Function to calculate MAPE
mape <- function(actual, predicted) {
  mean(abs((actual - predicted) / actual), na.rm = TRUE) * 100
}

# Function to safely calculate MAPE
mape <- function(actual, predicted) {
  valid_indices <- actual != 0  # Ignore zero values in y_test
  if (sum(valid_indices) == 0) {
    return(NA)  # If all values are zero, return NA instead of Inf
  }
  mean(abs((actual[valid_indices] - predicted[valid_indices]) / actual[valid_indices]), na.rm = TRUE) * 100
}

y_test <- test_set[[target_var]]

# Initialize results dataframe
results <- data.frame(Model = character(), RMSE = numeric(), R2 = numeric(), MAPE = numeric())

# Evaluate each model
for (model_name in names(models)) {   
  pred <- predict(models[[model_name]], newdata = test_set)
  
  rmse <- RMSE(pred, y_test)
  r2 <- R2(pred, y_test)
  mape_value <- mape(y_test, pred)  # Fixed MAPE function

  # If MAPE is Inf or NA, replace with a message
  if (is.infinite(mape_value) | is.na(mape_value)) {
    mape_value <- NA
  }
  
  # Append results
  results <- rbind(results, data.frame(Model = model_name, RMSE = rmse, R2 = r2, MAPE = mape_value))
  
  # Print performance metrics for each model
  cat("\n", model_name, "Model Performance:\n")
  cat("RMSE:", rmse, "\n")
  cat("R-squared:", r2, "\n")
  if (!is.na(mape_value)) {
    cat("MAPE:", mape_value, "%\n")
  } else {
    cat("MAPE: NA (due to zero values in y_test)\n")
  }
}

# Print overall model performance table
print(results)

```
## Feature Importance & AIC-based Model Selection
```{r}
# Feature Importance using caret
importance <- varImp(models$lm, scale = FALSE)
print(importance)
plot(importance)

# Train a standard linear regression model outside of caret
lm_model <- lm(as.formula(paste(target_var, "~ .")), data = train_set)

# Apply stepwise AIC selection
stepwise_model <- stepAIC(lm_model, direction = "both")

# Display results
summary(stepwise_model)

# Calculate VIF on the final model after stepwise selection
vif_values <- vif(stepwise_model)
print(vif_values)

# Remove features with VIF > 5 (indicating high multicollinearity)
high_vif_vars <- names(vif_values[vif_values > 5])
```
# Time Series Attempts

```{r redo data}
# Import datasets
setwd("/Users/madelinelee/Downloads/")

transactional_data <- read.csv("transactional_data.csv", stringsAsFactors = FALSE)
customer_profile <- read.csv("customer_profile.csv", stringsAsFactors = FALSE)
customer_address_and_zip_mapping <- read.csv("customer_address_and_zip_mapping.csv", stringsAsFactors = FALSE)
delivery_cost_data <- readxl::read_excel("delivery_cost_data.xlsx")

# Join transactional and customer profile data
transaction_customer_data <- transactional_data %>%
                             left_join(customer_profile, by = c("CUSTOMER_NUMBER" = "CUSTOMER_NUMBER"))

# Join customer address data
transaction_customer_data <- transaction_customer_data %>%
                             left_join(customer_address_and_zip_mapping, by = c("ZIP_CODE" = "zip"))

# Convert TRANSACTION_DATE to Date format
transaction_customer_data$TRANSACTION_DATE <- as.character(transaction_customer_data$TRANSACTION_DATE)
transaction_customer_data$Date <- as.Date(transaction_customer_data$TRANSACTION_DATE, format = "%m/%d/%Y")

# Ensure date conversion was successful
if (sum(is.na(transaction_customer_data$Date)) > 0) {
  warning("⚠️ WARNING: Some TRANSACTION_DATE values failed to convert.")
}

```

```{r aggregate sales}
# Aggregate daily total orders (calculated as ORDERED_CASES + ORDERED_GALLONS)
daily_orders <- transaction_customer_data %>%
  group_by(Date) %>%
  summarise(
    Total_Orders = sum(ORDERED_CASES, na.rm = TRUE) + sum(ORDERED_GALLONS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Date)

# Fill missing dates with 0 orders
daily_orders <- daily_orders %>%
  complete(Date = seq(min(Date), max(Date), by = "day"), fill = list(Total_Orders = 0))

# Split into train and test sets (80% train, 20% test)
split_point <- floor(0.8 * nrow(daily_orders))
train_ts <- daily_orders[1:split_point, ]
test_ts <- daily_orders[(split_point + 1):nrow(daily_orders), ]

```


## Train Models

### ARIMA
```{r auto ARIMA}
# Train ARIMA model
arima_model <- auto.arima(train_ts$Total_Orders, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)

# Forecast
arima_preds <- forecast(arima_model, h = nrow(test_ts))$mean
```

### ETS

```{r ETS}
# Train ETS model
ets_model <- ets(train_ts$Total_Orders)

# Forecast
ets_preds <- forecast(ets_model, h = nrow(test_ts))$mean
```


### Prophet
```{r prophet}
# Prepare Prophet data
prophet_data <- train_ts %>%
  rename(ds = Date, y = Total_Orders)

# Train Prophet model
prophet_model <- prophet(prophet_data)

# Create future dates
future_dates <- make_future_dataframe(prophet_model, periods = nrow(test_ts), freq = "day")

# Predict
prophet_forecast <- predict(prophet_model, future_dates)
prophet_preds <- tail(prophet_forecast$yhat, nrow(test_ts))

```

### XGBoost - Time Series

```{r xgboost}

# Feature Engineering
train_ts <- train_ts %>%
  tk_augment_timeseries_signature() %>%
  select(-diff, -year.iso, -month.lbl)

test_ts <- test_ts %>%
  tk_augment_timeseries_signature() %>%
  select(-diff, -year.iso, -month.lbl)

# Convert character/factor columns to numeric (one-hot encoding if needed)
train_ts <- train_ts %>%
  mutate(across(where(is.character), as.numeric, .names = "num_{col}")) %>%  # Convert characters
  mutate(across(where(is.factor), as.numeric))  # Convert factors

test_ts <- test_ts %>%
  mutate(across(where(is.character), as.numeric, .names = "num_{col}")) %>%  
  mutate(across(where(is.factor), as.numeric))



# Verify no character columns remain
str(train_ts)

train_ts[is.na(train_ts)] <- 0
test_ts[is.na(test_ts)] <- 0

# Convert train and test data to matrices
train_matrix <- as.matrix(train_ts %>% select(-Date, -Total_Orders))
test_matrix <- as.matrix(test_ts %>% select(-Date, -Total_Orders))

# Ensure data is numeric
if (!is.numeric(train_matrix)) stop("❌ ERROR: Train matrix contains non-numeric values!")

# Create XGBoost DMatrix
dtrain <- xgb.DMatrix(data = train_matrix, label = train_ts$Total_Orders)
dtest  <- xgb.DMatrix(data = test_matrix, label = test_ts$Total_Orders)



# Convert to XGBoost format
dtrain <- xgb.DMatrix(data = as.matrix(train_ts %>% select(-Date, -Total_Orders)), label = train_ts$Total_Orders)
dtest  <- xgb.DMatrix(data = as.matrix(test_ts %>% select(-Date, -Total_Orders)), label = test_ts$Total_Orders)

# Define XGBoost hyperparameter grid
best_xgb_model <- suppressWarnings(
  train(
    x = as.matrix(train_ts %>% select(-Date, -Total_Orders)),
    y = train_ts$Total_Orders,
    method = "xgbTree",
    tuneGrid = xgb_grid,
    trControl = trainControl(method = "cv", number = 5)
  )
)

# Train XGBoost model with tuning
best_xgb_model <- train(
  x = as.matrix(train_ts %>% select(-Date, -Total_Orders)),
  y = train_ts$Total_Orders,
  method = "xgbTree",
  tuneGrid = xgb_grid,
  trControl = trainControl(method = "cv", number = 5)
)

# Predict using XGBoost
xgb_preds <- predict(best_xgb_model, newdata = test_ts)

```

```{r evaluate}
# Function to calculate MAPE safely
mape <- function(actual, predicted) {
  valid_indices <- actual != 0  # Ignore zero values in y_test
  if (sum(valid_indices) == 0) {
    return(NA)  # If all values are zero, return NA instead of Inf
  }
  mean(abs((actual[valid_indices] - predicted[valid_indices]) / actual[valid_indices]), na.rm = TRUE) * 100
}

# Evaluate Models
models_results <- data.frame(Model = character(), RMSE = numeric(), MAPE = numeric(), MAE = numeric())

# ARIMA Evaluation
models_results <- rbind(models_results, data.frame(
  Model = "ARIMA",
  RMSE = RMSE(arima_preds, test_ts$Total_Orders),
  MAPE = mape(test_ts$Total_Orders, arima_preds),
  MAE = MAE(arima_preds, test_ts$Total_Orders)
))

# ETS Evaluation
models_results <- rbind(models_results, data.frame(
  Model = "ETS",
  RMSE = RMSE(ets_preds, test_ts$Total_Orders),
  MAPE = mape(test_ts$Total_Orders, ets_preds),
  MAE = MAE(ets_preds, test_ts$Total_Orders)
))

# Prophet Evaluation
models_results <- rbind(models_results, data.frame(
  Model = "Prophet",
  RMSE = RMSE(prophet_preds, test_ts$Total_Orders),
  MAPE = mape(test_ts$Total_Orders, prophet_preds),
  MAE = MAE(prophet_preds, test_ts$Total_Orders)
))

# XGBoost Evaluation
models_results <- rbind(models_results, data.frame(
  Model = "XGBoost",
  RMSE = RMSE(xgb_preds, test_ts$Total_Orders),
  MAPE = mape(test_ts$Total_Orders, xgb_preds),
  MAE = MAE(xgb_preds, test_ts$Total_Orders)
))

# Print model performance
print(models_results)

```


