---
title: "Swire EDA"
author: "Maddie Lee"
date: "2025-02-10"
output: html_document
---
# Business Statement


# Exploration Tasks
```{r, message=FALSE, warning = FALSE}
# Load libraries
library(tidyverse)
library(readr) 
library(readxl)
library(dplyr)
library(DataExplorer) 
library(ggplot2)
library(caret)
library(skimr)
library(tidymodels)
library(patchwork)
library(missForest)
library(kableExtra)
library(ROSE)
library(car)
library(ggcorrplot)
```

## Cleaning the Data

### How Does the Initial Data Look?
```{r read data}

# Read CSV files
customer_profile <- read_csv("~/Downloads/customer_profile.csv")
customer_address <- read_csv("~/Downloads/customer_address_and_zip_mapping.csv")
transactional_data <- read_csv("~/Downloads/transactional_data.csv")

# Read Excel file
delivery_cost_data <- read_excel("~/Downloads/delivery_cost_data.xlsx")
```

```{r view data}
head(customer_profile)
head(customer_address)
head(transactional_data)
head(delivery_cost_data)
```

Consider splitting customer address into city or state?

```{r chr to factor}
# Check for duplicated data
anyDuplicated(customer_profile)
anyDuplicated(customer_address)
anyDuplicated(transactional_data)
anyDuplicated(delivery_cost_data)

# Check data type of variables
str(customer_profile)
str(customer_address)
str(transactional_data)
str(delivery_cost_data)

# Convert 
convert_to_factor <- function(df) {
  df[] <- lapply(df, function(x) if (is.character(x)) as.factor(x) else x)
  return(df)
}

# Apply the function to all datasets
customer_address <- convert_to_factor(customer_address)
customer_profile <- convert_to_factor(customer_profile)
transactional_data <- convert_to_factor(transactional_data)
delivery_cost_data <- convert_to_factor(delivery_cost_data)

# Convert zipcode, customer number, primary group number, week, and year to factor

# Function to convert specific columns to factors
convert_to_factor <- function(df, columns) {
  df <- df %>%
    mutate(across(all_of(columns), as.factor))
  return(df)
}

# Convert in customer profile
customer_profile <- convert_to_factor(customer_profile, c("ZIP_CODE", "CUSTOMER_NUMBER", "PRIMARY_GROUP_NUMBER"))

# Convert in transactional data
transactional_data <- convert_to_factor(transactional_data, c("CUSTOMER_NUMBER",  "WEEK", "YEAR"))

# Convert in customer address dataset 
customer_address <- convert_to_factor(customer_address, c("zip"))


# Verify conversion
str(customer_address)
str(customer_profile)
str(transactional_data)
str(delivery_cost_data)
```

```{r numeric variable distributions}
# Function to plot numeric variable distributions
plot_numeric_distributions <- function(df, dataset_name) {
  numeric_cols <- df %>% select(where(is.numeric)) %>% names()
  
  if (length(numeric_cols) > 0) {
    for (col in numeric_cols) {
      p <- ggplot(df, aes(x = .data[[col]])) +
        geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.7) +
        geom_density(aes(y = ..count..), color = "red", size = 1) +
        labs(title = paste("Distribution of", col, "in", dataset_name),
             x = col,
             y = "Frequency") +
        theme_minimal()
      
      print(p)  # Display plot
    }
  } else {
    cat("\nNo numeric variables found in", dataset_name, "\n")
  }
}

# Generate distribution plots for each dataset
plot_numeric_distributions(customer_address, "Customer Address and ZIP Mapping")
plot_numeric_distributions(customer_profile, "Customer Profile")
plot_numeric_distributions(transactional_data, "Transactional Data")
plot_numeric_distributions(delivery_cost_data, "Delivery Cost Data")

summary(transactional_data$DELIVERED_CASES)
summary(transactional_data$DELIVERED_GALLONS)
summary(transactional_data$ORDERED_CASES)
summary(transactional_data$ORDERED_GALLONS)
summary(transactional_data$LOADED_CASES)
summary(transactional_data$LOADED_GALLONS)
```
Negative values of delivered_gallons and delivered cases?

Different values of ordered, delivered, and loaded in cases and gallons


### Are there Columns with Near-Zero Variance?

```{r near zero variance}
# View variables that are near zero variance 
nearZeroVar(customer_profile,saveMetrics = TRUE)
nearZeroVar(customer_profile)
# View variables that are near zero variance 
nearZeroVar(transactional_data,saveMetrics = TRUE)
nearZeroVar(transactional_data)
# View variables that are near zero variance 
nearZeroVar(delivery_cost_data,saveMetrics = TRUE)
nearZeroVar(delivery_cost_data)
```
No columns appear to have near zero variance 


### What is the scope of missing data?

```{r missing data}
# Function to check missing values per column
check_missing_values <- function(df, df_name) {
  missing_counts <- colSums(is.na(df))
  missing_percent <- round((missing_counts / nrow(df)) * 100, 2)
  missing_df <- data.frame(Column = names(df), Missing_Count = missing_counts, Missing_Percent = missing_percent)
  
  # Print missing values summary
  cat("\nMissing Values Summary for", df_name, "\n")
  print(missing_df %>% arrange(desc(Missing_Count)))  # Sort by most missing values
  
  return(missing_df)
}

# Apply function to all datasets
missing_customer_profile <- check_missing_values(customer_profile, "customer_profile")
missing_customer_address <- check_missing_values(customer_address, "customer_address")
missing_transactional_data <- check_missing_values(transactional_data, "transactional_data")
missing_delivery_cost <- check_missing_values(delivery_cost_data, "delivery_cost_data")

```


Largest missing values in Primary Group Number (almost 60%)


## Feature Engineering 

## Target Variable
